---
title: "Running Pv3Rs for Peru data"
author: "Shazia Ruybal-Pesántez"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide 
    code_download: true
    fig_width: 8
    fig_height: 6
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(janitor)
library(here)
library(patchwork)
library(reshape2)

# install.packages('Pv3Rs', repos = c('https://plasmogenepi.r-universe.dev', 'https://cloud.r-project.org'))

# devtools::install_github("aimeertaylor/Pv3Rs")
# Pv3Rs downloaded from Github on 7 Aug 2024
library(Pv3Rs)
```

## Set global variables
From TO/JS - This section sets some flags that can be tweaked to re-run the analysis with different filtering parameters:

`RUN_EXAMPLE`: allows discarding alleles that have a frequency below a threshold (defaults to FALSE),
`WITHIN_INDIVIDUAL_ALLELE_FREQ_THR`: allows discarding alleles that have a frequency below a threshold (defaults to 0),
`BENCHMARK_MARKERS`: list of markers to include in the analysis (defaults to the full list established by Jason, could be reduced down to 3 for memory optimization),
`MAX_MOI_TO_INCLUDE`: allows discarding participant data when they exhibit very complex infection patterns, counted as the total number of clones across all infection events (defaults to 8),
`PRIOR_3RS`: a named vector with prior probabilities of recrudescence (C), relapse (L) or reinfection (I) (defaults to uniform, i.e. 1/3 each, but tweaked to minimize recrudescence in the present study).

```{r global vars}
## RUN_EXAMPLE
##   Boolean flag to enable/disable running the example Pv3Rs posterior 
##   computation on a single episode sampled at random from the dataset.
## DEFAULT: FALSE
RUN_EXAMPLE <- TRUE

## WITHIN_INDIVIDUAL_ALLELE_FREQ_THR
##   Minimum threshold above which an allele is preserved in individual
##   haplotype data to be preserved when reconstructing infection 
##   history. Note that this will *not* discard these alleles from the
##   population-level allele frequency that is derived from the initial
##   visit, but only discard that allele from individual observations 
##   when it is 'too rare to be exploited'.
## DEFAULT: 0
WITHIN_INDIVIDUAL_ALLELE_FREQ_THR <- 0

## BENCHMARK_MARKERS
##   Vector of character string with names matching those from the 
##   markers of interest for amplicon sequencing. Only the markers 
##   listed in that vector will be preserved in individual-level data
##   and identity of infection will be solely based on the alleles 
##   observed for these markers.
##   The markers are listed by importance as discovered by Jason.
## DEFAULT: all markers (could be suboptimal/too memory-consuming?)
BENCHMARK_MARKERS <- c(
  "Chr05",
  "Chr07", 
  "Chr09", 
  "Chr10",
  "Chr08",
  "Chr13",
  "Chr11",
  "Chr03",
  "Chr01",
  "Chr02",
  "Chr14"
)

## MAX_MOI_TO_INCLUDE
##   As per Aimee's guidelines: 
##     We do not recommend running compute_posterior() for data 
##     whose total genotype count exceeds eight, where the total 
##     genotype count is the sum of per-episode maximum per-marker 
##     allele counts.
##   The MAX_MOI_TO_INCLUDE expects an integer that will discard all
##   individuals having a summed MOI > 8 across all recorded episodes.
## DEFAULT: 8
MAX_MOI_TO_INCLUDE <- 8

## PRIORS_3RS
##   A vector of probabilities, summing to 1, corresponding to 
##   the probability of each stage for the 3Rs for Pv episodes.
##   The vector order is re(C)rudescence, re(L)apse, re(I)nfection.
##   In this clinical trial, we assume recrudescence is possible so, 
##   we use the default priors 
## DEFAULT: c("C" = 1/3, "L" = 1/3, "I" = 1/3)
# PRIOR_3RS <- c("C" = 1/3, "L" = 1/3, "I" = 1/3)
# Note: the below probabilities were used for SeroTAT study because it does not involve treatment at baseline
# We will use the same for the Peru community survey
PRIOR_3RS <- c("C" = 0.10, "L" = 0.45, "I" = 0.45)
```

## Peru data curation

Read in data from Peru:
```{r read peru data}
peru <- read.csv(here("data/final", "merged_PE.csv"))

# head(peru)
# names(peru)
```

### Number of unique haplotypes (ie marker cardinality):
```{r n haps}
peru %>% 
  group_by(marker_id) %>% 
  summarise(unique_haplotypes = n_distinct(haplotype))
```

### Counts of each haplotype:
```{r hap counts}
peru %>% count(marker_id, haplotype)
```

### "MOI":
```{r moi}
peru %>% 
  group_by(PatientName, marker_id) %>%
  summarise(n_haplotypes = n(), .groups = 'drop') %>%
  group_by(PatientName) %>% 
  summarise(max_moi = max(n_haplotypes)) 
```

### Recurrences
Participants with more than one episode for inference (n=40 participants, n=87 Pv isolates):
```{r participant with recurrences}
peru %>% 
  select(PatientName, episodes) %>% 
  distinct() %>% 
  arrange(PatientName) %>% 
  filter(episodes>1) 

ids_recurrent <- peru %>% clean_names() %>% filter(episodes > 1) %>% distinct(patient_name)
peru_recurrent <- peru %>% clean_names() %>% filter(episodes > 1)

# setdiff(ids_recurrent$patient_name, peru_recurrent$patient_name)
# setdiff(peru_recurrent$patient_name, ids_recurrent$patient_name)

# peru_recurrent %>% distinct(sample_id) %>% arrange(sample_id) # 87 isolates
```

### Data on episode number and time since last episode
```{r episode summary}
episode_summary <- peru_recurrent %>% 
  distinct(patient_name, date, sample_id, day) %>% 
  mutate(date = mdy(date)) %>%
  arrange(patient_name, date) %>% 
  group_by(patient_name) %>% 
  mutate(
    # get the episode number
    episode_number = row_number(),
    # calculate days since enrolment, just a check with dates
    days_since_enrolment = as.integer(date - min(date[day == "Day 0"])),
    # calculate days since last episode using lag() and making enrolment episodes 0 days since last
    days_since_last_episode = replace_na(as.integer(date - lag(date)), 0)
  ) %>% 
  ungroup()
```

Note that in the Peru study this is not necessarily a person's "first" episode, rather the first episode in the evaluated study period
```{r plot recurrences, fig.height=8, fig.width=8}
episode_summary %>% 
  distinct(patient_name, sample_id, days_since_enrolment) %>%
  ggplot(aes(x = days_since_enrolment, y = reorder(patient_name, days_since_enrolment))) +
    geom_line(aes(group = patient_name), color = "darkgrey") +
    geom_point() +
    labs(x = "Days since first episode during study period",
         y = "Participant") +
    theme_bw() 
```

## Haplotype frequencies
We want to remove any haplotypes that appear only once at very low frequencies. Haplotypes should already be filtered to be observed in at least 2 samples and within-host frequency >=1% (in full dataset).

```{r singleton haps in recurrent dataset}
singleton_haps <- peru_recurrent %>% 
  select(sample_id, marker_id, haplotype, frequency, count) %>% 
  count(haplotype) %>% 
  arrange(n) %>% 
  filter(n==1) %>% 
  pull(haplotype)

peru_recurrent %>% 
  filter(haplotype %in% singleton_haps) %>% 
  ggplot(aes(x = haplotype, y = count)) +
    geom_hline(yintercept = 100, linetype = "dashed") +
    geom_point(aes(color = frequency, shape = day), size = 3) +
    labs(x = "singleton haplotype",
         y = "read count",
         color = "within-sample frequency",
         shape = "timepoint") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))

peru %>% 
  filter(haplotype %in% singleton_haps) %>% 
  ggplot(aes(x = haplotype, y = count)) +
    geom_hline(yintercept = 100, linetype = "dashed") +
    geom_point(aes(color = frequency, shape = day), size = 3) +
    labs(x = "singleton haplotype",
         y = "read count",
         color = "within-sample frequency",
         shape = "timepoint") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```
## Haplotypes for inference
For Peru, most recurrences would be expected to be either reinfections or relapses, given that this study was an observational community cohort in two sites in Peru. As per Aimee's suggestion, we can estimate population-level allele frequencies based on all episodes in our dataset (we will use the entire dataset (n=93 participants), not just those the subset that experienced recurrences (n=40). This should be unbiased as both reinfections and relapses are draws from the mosquito population (relapses would be time-lagged draws) in the absence of any systematic within-patient selection of recrudescent parasites.
```{r haps for inference}
# n=1530 haps
all_haps <- peru %>% 
  pull(haplotype)

peru %>%
  select(sample_id, marker_id, haplotype, frequency, count) %>%
  arrange(frequency, count)
```

## Final data for inference
```{r final analysis data}
analysis_data <- peru_recurrent %>%
  select(subject_id = patient_name, 
         sample_id, 
         community = com_y, 
         timepoint = day,
         visit_date = date,
         age_years = age, 
         sex, 
         episodes, 
         marker_id, 
         haplotype, 
         frequency,
         mean_moi,
         max_moi) %>% 
  # add extra epi info on episode number and time since last episode
  left_join(episode_summary %>% select(subject_id = patient_name, 
                                       sample_id, 
                                       episode_number,
                                       days_since_enrolment,
                                       days_since_last_episode),
            by = c("subject_id", "sample_id")) %>% 
  # ensure dates are date class
  mutate(visit_date = mdy(visit_date))
```

### Population-level allele frequencies
```{r test allele freqs, eval=F}
test_df <- peru %>% 
  # filter data frame by marker
  filter(marker_id == "Chr05")

test_df %>% 
      mutate(haplotype = factor(haplotype)) %>% 
      select(sample_id, haplotype, frequency) %>% 
      pivot_wider(names_from = haplotype, 
                  values_from = frequency, 
                  values_fill = 0) %>% 
      pivot_longer(cols = -sample_id, 
                   names_to = "haplotype", 
                   values_to = "frequency") %>% 
      # Get population-level haplotype frequency, 
      # correcting for when within-individual sum is not equal 
      # to 1, as can happen when a minority clone is <2% 
      group_by(sample_id) %>%
      mutate(frequency = frequency / sum(frequency, na.rm = TRUE)) %>%
  
      # get population-level mean freq
      group_by(haplotype) %>% 
      summarise(poplevel_mean_freq = mean(frequency, na.rm = TRUE)) %>% 
      adorn_totals("row")
```

```{r pop level allele freqs}
## Derive pop-level allele frequency - This is modified from Thomas/Jason script for our data

# Use data from the entire dataset
fs <- peru %>% 
  
  # split data frame by marker
  group_by(marker_id) %>% 
  group_split() %>% 
  
  # Derive a within-marker list of frequencies, by individual
  lapply(function(x) {
    x <- x %>% 
      # Build a within-individual frequency table that 
      # always includes every haplotype (even the ones absent)
      mutate(haplotype = factor(haplotype)) %>% 
      select(sample_id, haplotype, frequency) %>% 
      pivot_wider(names_from = haplotype, 
                  values_from = frequency, 
                  values_fill = 0) %>% 
      pivot_longer(cols = -sample_id, 
                   names_to = "haplotype", 
                   values_to = "frequency") %>% 
      # Get population-level haplotype frequency, 
      # correcting for when within-individual sum is not equal 
      # to 1, as can happen when a minority clone is <2%
      group_by(sample_id) %>% 
      mutate(frequency = frequency / sum(frequency, na.rm = TRUE)) %>% 
      group_by(haplotype) %>% 
      summarise(frequency_pop_mean = mean(frequency, na.rm = TRUE))
    
    return(deframe(x))
  }) %>% 
  setNames(nm = peru %>% group_by(marker_id) %>% group_keys() %>% pull(marker_id))

# Here we can also save as dataframe for easier printing and table-ready for paper
fs_df <- peru %>%

  # Group by marker_id and sample_id for further calculations
  group_by(marker_id, sample_id) %>%

  # Build a within-individual frequency table that always includes every haplotype (even the ones absent)
  mutate(haplotype = factor(haplotype)) %>%
  select(marker_id, sample_id, haplotype, frequency) %>%
  pivot_wider(names_from = haplotype, values_from = frequency, values_fill = list(frequency = 0)) %>%
  pivot_longer(cols = -c(marker_id, sample_id), names_to = "haplotype", values_to = "frequency") %>%

  # Get population-level haplotype frequency, correcting for when within-individual sum is not equal to 1
  group_by(marker_id, sample_id) %>%
  mutate(frequency = frequency / sum(frequency, na.rm = TRUE)) %>%
  group_by(marker_id, haplotype) %>%
  summarise(frequency_pop_mean = mean(frequency, na.rm = TRUE), .groups = 'drop') %>%

  # Ensure haplotypes are correctly associated with their marker_id - note that this works for us because , in future would have to make this flexible to allow for haplotype names that are not reliant on having marker_id
  filter(str_detect(haplotype, marker_id))

fs_df
```

## Pv3Rs
### Example for one participant: M1A020B 
```{r pv3rs one example}
# This has been modified from Thomas/Jason script 

## Pick an individual at random to run Pv3Rs
# indiv_name <- sample(unique(analysis_data$subject_id), size = 1)
indiv_name <- "M1A020B"

## Prepare the data
# 1- Subset haplotype data to specific individual and apply filters
indiv_haplotype_data <- analysis_data %>% 
  # Restrict to a single patient
  filter(subject_id == indiv_name) %>% 
  # Restrict to a subset of markers, if needed
  filter(marker_id %in% BENCHMARK_MARKERS) %>% 
  # Restrict to summed MOI below threshold
  group_by(subject_id, episode_number, marker_id) %>% 
  mutate(MOI_per_marker = sum(n())) %>% 
  group_by(subject_id, episode_number) %>% 
  mutate(MOI_per_episode = max(MOI_per_marker, na.rm = TRUE)) %>% 
  ungroup() %>% 
  mutate(marker_id = factor(marker_id, levels = BENCHMARK_MARKERS))

# 2- Calculate per-episode and per-participant MOI for PvR3S eligibility
indiv_MOI <- indiv_haplotype_data %>% 
  select(subject_id, episode_number, 
         marker_id, starts_with("MOI_")) %>% 
  distinct() %>% 
  group_by(subject_id, episode_number) %>% 
  # Get highest per-marker MOI only for each episode
  # (drop marker_id in case of ties with highest per-marker MOI)
  select(-marker_id) %>% 
  distinct() %>% 
  filter(MOI_per_marker == max(MOI_per_marker, na.rm = TRUE)) %>% 
  ungroup() %>% 
  mutate(MOI_summed = sum(MOI_per_episode, na.rm = TRUE))
```

Prepare test data
```{r}
# Finish data preparation
  indiv_haplotype_data <- indiv_haplotype_data %>% 
    group_by(episode_number) %>% 
    group_split() %>% 
    lapply(function(x) {
      res <- x %>% 
        select(sample_id, episode_number, 
               marker_id, haplotype, frequency) %>% 
        # For sensitivity analysis, allow to include/drop allele 
        # based on their within-individual frequency
        filter(frequency >= WITHIN_INDIVIDUAL_ALLELE_FREQ_THR) %>% 
        select(-sample_id, -episode_number, -frequency) %>% 
        distinct() %>% 
        # Prevent dropping of markers that are not characterised 
        # by setting .drop to FALSE
        group_by(marker_id, .drop = FALSE) %>% 
        group_split() %>% 
        lapply(function(y) {
          unique(y$haplotype)
        })
      
      # Returned a list named with each episode, 
      # setting marker allele to NA in case none are observed
      return(lapply(setNames(res, BENCHMARK_MARKERS), 
                    function(y) {
                      if (length(y) == 0) return(NA) else return(y)
                    }))
    })
```

```{r run compute post one indiv}
# Run Aimee's posterior estimation
indiv_posterior <- Pv3Rs::compute_posterior(y  = indiv_haplotype_data, 
                                            fs = fs[BENCHMARK_MARKERS])

indiv_posterior
```
The joint posterior probability of reinfection is 99.8%, when we look at the genetic data, this is in line with the inferred probability of relapse. 
```{r plot haps}
peru_recurrent %>% 
  filter(patient_name == "M1A020B") %>% 
  ggplot(aes(x = haplotype, y = factor(date), fill = factor(haplotype))) +
  geom_tile() +
    facet_grid(~marker_id, scales = "free_x") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

### Run Pv3Rs one random individual
```{r subset haplotype data at random}
# This has been modified from Thomas/Jason script 

## Pick an individual at random to run Pv3Rs
indiv_name <- sample(unique(analysis_data$subject_id), size = 1)

## Prepare the data
# 1- Subset haplotype data to specific individual and apply filters
indiv_haplotype_data <- analysis_data %>% 
  # Restrict to a single patient
  filter(subject_id == indiv_name) %>% 
  # Restrict to a subset of markers, if needed
  filter(marker_id %in% BENCHMARK_MARKERS) %>% 
  # Restrict to summed MOI below threshold
  group_by(subject_id, episode_number, marker_id) %>% 
  mutate(MOI_per_marker = sum(n())) %>% 
  group_by(subject_id, episode_number) %>% 
  mutate(MOI_per_episode = max(MOI_per_marker, na.rm = TRUE)) %>% 
  ungroup() %>% 
  mutate(marker_id = factor(marker_id, levels = BENCHMARK_MARKERS))

# 2- Calculate per-episode and per-participant MOI for PvR3S eligibility
indiv_MOI <- indiv_haplotype_data %>% 
  select(subject_id, episode_number, 
         marker_id, starts_with("MOI_")) %>% 
  distinct() %>% 
  group_by(subject_id, episode_number) %>% 
  # Get highest per-marker MOI only for each episode
  # (drop marker_id in case of ties with highest per-marker MOI)
  select(-marker_id) %>% 
  distinct() %>% 
  filter(MOI_per_marker == max(MOI_per_marker, na.rm = TRUE)) %>% 
  ungroup() %>% 
  mutate(MOI_summed = sum(MOI_per_episode, na.rm = TRUE))
```


Now that the data has been subset, Pv3Rs::compute_posterior() will be called (only if the participant meets eligibility criteria as defined by the global variables set earlier in this report).

```{r run Pv3Rs one random indiv, eval=F}
if (RUN_EXAMPLE & unique(indiv_MOI$MOI_summed) <= MAX_MOI_TO_INCLUDE) {
  # Verbose
  cat("Running Pv3Rs for: ", 
      indiv_name, 
      " (summed MOI = ", 
      unique(indiv_MOI$MOI_summed), 
      ").\n", 
      sep = "")
  
  # Finish data preparation
  indiv_haplotype_data <- indiv_haplotype_data %>% 
    group_by(episode_number) %>% 
    group_split() %>% 
    lapply(function(x) {
      res <- x %>% 
        select(sample_id, episode_number, 
               marker_id, haplotype, frequency) %>% 
        # For sensitivity analysis, allow to include/drop allele 
        # based on their within-individual frequency
        filter(frequency >= WITHIN_INDIVIDUAL_ALLELE_FREQ_THR) %>% 
        select(-sample_id, -episode_number, -frequency) %>% 
        distinct() %>% 
        # Prevent dropping of markers that are not characterised 
        # by setting .drop to FALSE
        group_by(marker_id, .drop = FALSE) %>% 
        group_split() %>% 
        lapply(function(y) {
          unique(y$haplotype)
        })
      
      # Returned a list named with each episode, 
      # setting marker allele to NA in case none are observed
      return(lapply(setNames(res, BENCHMARK_MARKERS), 
                    function(y) {
                      if (length(y) == 0) return(NA) else return(y)
                    }))
    })
  
  # Run Aimee's posterior estimation
  indiv_posterior <- Pv3Rs::compute_posterior(y  = indiv_haplotype_data, 
                                              fs = fs[BENCHMARK_MARKERS])
} else {
  # Verbose
  cat("NOT Running Pv3Rs for: ", 
      indiv_name, 
      " because either summed MOI exceeds threshold (observed = ", 
      unique(indiv_MOI$MOI_summed), 
      ", MAX_MOI_TO_INCLUDE = ", 
      MAX_MOI_TO_INCLUDE, 
      "), or RUN_EXAMPLE was set to FALSE.\n", 
      sep = "")
  
  # Return NULL
  indiv_posterior <- NULL
}

indiv_posterior
```
```{r plot haps for indiv, eval=F}
peru_recurrent %>% 
  filter(patient_name == indiv_name) %>% 
  ggplot(aes(x = haplotype, y = factor(date), fill = factor(haplotype))) +
  geom_tile() +
    facet_grid(~marker_id, scales = "free_x") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

### Run for all participants
```{r run pv3rs for all}
# Start timer
t_start <- Sys.time()

# Initialize an empty list to store the results
indiv_posteriors <- list()

# Loop through each unique individual name
for (indiv_name in unique(analysis_data$subject_id)) {
  # Verbose
  cat("ID : ", indiv_name, "...\n", sep = "")
  
  ## Prepare the data
  # 1- Subset haplotype data to specific individual and apply filters
  indiv_haplotype_data <- analysis_data %>% 
    # Restrict to a single patient
    filter(subject_id == indiv_name) %>% 
    # Restrict to a subset of markers
    filter(marker_id %in% BENCHMARK_MARKERS) %>% 
    # Restrict to summed MOI below threshold
    group_by(subject_id, episode_number, marker_id) %>% 
    mutate(MOI_per_marker = sum(n())) %>% 
    group_by(subject_id, episode_number) %>% 
    mutate(MOI_per_episode = max(MOI_per_marker, na.rm = TRUE)) %>% 
    ungroup() %>% 
    mutate(marker_id = factor(marker_id, levels = BENCHMARK_MARKERS))
  
  # 2- Calculate per-episode and per-participant MOI for PvR3S eligibility
  indiv_MOI <- indiv_haplotype_data %>% 
    select(subject_id, episode_number, 
           marker_id, starts_with("MOI_")) %>% 
    distinct() %>% 
    group_by(subject_id, episode_number) %>% 
    # Get highest per-marker MOI only for each episode
    # (drop marker_id in case of ties with highest per-marker MOI)
    select(-marker_id) %>% 
    distinct() %>% 
    filter(MOI_per_marker == max(MOI_per_marker, na.rm = TRUE)) %>% 
    ungroup() %>% 
    mutate(MOI_summed = sum(MOI_per_episode, na.rm = TRUE))
  
  ## Run Pv3Rs only if MOI below threshold         
  if (unique(indiv_MOI$MOI_summed) <= MAX_MOI_TO_INCLUDE) {
    # Verbose
    cat("Running Pv3Rs for: ", 
        indiv_name, 
        " (summed MOI = ", 
        unique(indiv_MOI$MOI_summed), 
        ").\n", 
        sep = "")
    
    # Preserve rounds for naming the output list
    indiv_episode <- unique(indiv_haplotype_data$episode_number)
    
    # Finish data preparation
    indiv_haplotype_data <- indiv_haplotype_data %>% 
      group_by(episode_number) %>% 
      group_split() %>% 
      lapply(function(x) {
        res <- x %>% 
          select(sample_id, episode_number, 
                 marker_id, haplotype, frequency) %>% 
          # For sensitivity analysis, allow to include/drop allele 
          # based on their within-individual frequency
          filter(frequency >= WITHIN_INDIVIDUAL_ALLELE_FREQ_THR) %>% 
          select(-sample_id, -episode_number, -frequency) %>% 
          distinct() %>% 
          # Prevent dropping of markers that are not characterized 
          # by setting .drop to FALSE
          group_by(marker_id, .drop = FALSE) %>% 
          group_split() %>% 
          lapply(function(y) {
            unique(y$haplotype)
          })
        
        # Returned a list named with each episode, 
        # setting marker allele to NA in case none are observed
        return(lapply(setNames(res, BENCHMARK_MARKERS), 
                      function(y) {
                        if (length(y) == 0) return(NA) else return(y)
                      }))
      })
    
    # Run Aimee's posterior estimation
    indiv_posterior <- compute_posterior(y     = indiv_haplotype_data,
                                         fs    = fs[BENCHMARK_MARKERS], 
                                         prior = matrix(PRIOR_3RS, 
                                                        nrow     = length(indiv_haplotype_data), 
                                                        ncol     = length(PRIOR_3RS), 
                                                        byrow    = TRUE, 
                                                        dimnames = list(c(1:length(indiv_haplotype_data)), 
                                                                        names(PRIOR_3RS))))
    
  } else {
    # Verbose
    cat("NOT Running Pv3Rs for: ", 
        indiv_name, 
        " because summed MOI exceeds threshold (observed = ", 
        unique(indiv_MOI$MOI_summed), 
        ", MAX_MOI_TO_INCLUDE = ", 
        MAX_MOI_TO_INCLUDE, 
        ").\n", 
        sep = "")
    
    # Return NULL
    indiv_episode    <- unique(indiv_haplotype_data$episode_number)
    indiv_posterior <- NULL
  }
  
  # Append the results to the list
  indiv_posteriors[[indiv_name]] <- list("subject_id" = indiv_name, 
                                         "episode_number"       = indiv_episode, 
                                         "Pv3Rs"       = indiv_posterior)
}

# End timer
t_end <- Sys.time()
cat("Pv3Rs for the whole dataset took : ", as.numeric(difftime(time1 = t_end, 
                                                               time2 = t_start, 
                                                               units = "secs"))/60, " mins", "\n", sep = "")

# Present the marginal data in a clearer format
indiv_posteriors_marginal <- do.call(rbind, 
                                     lapply(indiv_posteriors, function(x) {
                                       if (!is.null(x[["Pv3Rs"]])) {
                                         return(data.frame("subject_id"               = x[["subject_id"]], 
                                                           "episode_number"                     = x[["episode_number"]][-1], 
                                                           "Posterior_marginal_prob_C" = x[["Pv3Rs"]]$marg[, "C"], 
                                                           "Posterior_marginal_prob_L" = x[["Pv3Rs"]]$marg[, "L"], 
                                                           "Posterior_marginal_prob_I" = x[["Pv3Rs"]]$marg[, "I"]))
                                       } else {
                                         return(NULL)
                                       }
                                       
                                     }))
row.names(indiv_posteriors_marginal) <- 1:nrow(indiv_posteriors_marginal)

# Present the joint posterior estimates in a clearer format
indiv_posteriors_joint <- do.call(rbind, 
                                  lapply(indiv_posteriors, function(x) {
                                    if (!is.null(x[["Pv3Rs"]])) {
                                      joint_probs <- x[["Pv3Rs"]]$joint
                                      # Extract the state pairs and probabilities
                                      state_pairs <- names(joint_probs)
                                      prob_values <- as.numeric(joint_probs)
                                      
                                      # Create a data frame with subject_id, episode_number, and joint probabilities
                                      return(data.frame("subject_id"      = rep(x[["subject_id"]], length(state_pairs)), 
                                                        "episode_number"  = rep(x[["episode_number"]][-1], each = length(state_pairs)),
                                                        "state_pair"      = state_pairs, 
                                                        "joint_probability" = prob_values))
                                    } else {
                                      return(NULL)
                                    }
                                  }))
row.names(indiv_posteriors_joint) <- 1:nrow(indiv_posteriors_joint)

# Save Pv3Rs output because it's time consuming and 
# we don't want to re-run it every time.
save(list = c("RUN_EXAMPLE", "WITHIN_INDIVIDUAL_ALLELE_FREQ_THR", "BENCHMARK_MARKERS", "MAX_MOI_TO_INCLUDE", "PRIOR_3RS", 
              "analysis_data", "fs", 
              "indiv_posteriors", "indiv_posteriors_marginal", "indiv_posteriors_joint"), 
     file = paste0("./outputs/Pv3Rs_peru_posteriors_", 
                   strftime(Sys.time(), format = "%Y%m%d_%H%M%S"), 
                   ".RData"))
```
### Explore the posterior
The marginal probabilities give us the probability of the three states for each recurrent episode. However, this does not consider the joint probability of different states when a person experienced more than one recurrent episode. 

```{r marginal probs}
indiv_posteriors_marginal
```

The joint probabilites give us values for each possible combination of states, depending on the number of recurrent episodes experienced by the participant. 
```{r joint probs}
joint_summary <- indiv_posteriors_joint %>% 
                    group_by(subject_id, state_pair, joint_probability) %>% 
                    filter(episode_number == max(episode_number)) %>% 
                    mutate(percentage = round(joint_probability*100, 3),
                           total_recurrences = episode_number-1) %>%
                    select(-episode_number) %>% 
                    arrange(subject_id, total_recurrences, percentage) %>% 
                    relocate(total_recurrences, .before = state_pair)

joint_summary
```

## Plot marginal probabilities
```{r marg probs}
marginal_summary <- indiv_posteriors_marginal %>% 
  pivot_longer(cols = !subject_id & !episode_number, 
               names_to = "posterior_type", 
               values_to = "posterior_value") %>% 
  mutate(posterior_classification = case_when(posterior_type == "Posterior_marginal_prob_C" ~ "Recrudescence",
                                              posterior_type == "Posterior_marginal_prob_L" ~ "Relapse",
                                              posterior_type == "Posterior_marginal_prob_I" ~ "Reinfection"),
         posterior_classification = factor(posterior_classification, 
                                           levels = c("Relapse", "Recrudescence", "Reinfection"))) %>% 
  select(-posterior_type)

marginal_summary
```

### By episode number
```{r plot probabilities, fig.height=8, fig.width=8}
marginal_summary %>% 
  group_by(subject_id) %>%
  arrange(desc(posterior_value), .by_group = TRUE) %>%
  mutate(subject_id = factor(subject_id, levels = unique(subject_id[order(posterior_value, decreasing = TRUE)]))) %>%
  
    ggplot(aes(x = factor(episode_number), y = posterior_value, group = episode_number, fill = posterior_classification)) + 
    geom_bar(stat = "identity", position = "fill") +
    scale_fill_manual(values = c("Relapse" = "turquoise3",
                                   "Recrudescence" = "skyblue4",
                                   "Reinfection" = "magenta3")) +
    
    scale_x_discrete(expand = c(0, 0)) +
    scale_y_continuous(expand = c(0, 0)) +
    
    labs(x     = "Episode number", 
         y     = "Posterior probability", 
         fill = "") + 
    theme_bw() +
    facet_wrap(~subject_id)
```
### By community
```{r plot probabilites by community, fig.height=8, fig.width=8}
marginal_summary %>% 
  left_join(analysis_data %>% select(subject_id, community) %>% distinct(), 
            by = "subject_id") %>% 
  group_by(subject_id) %>% 
  arrange(desc(posterior_value), .by_group = TRUE) %>% 
  mutate(subject_id = factor(subject_id, levels = unique(subject_id[order(posterior_value, decreasing = TRUE)]))) %>% 
  ggplot(aes(x = factor(episode_number), y = posterior_value, group = episode_number, fill = posterior_classification)) + 
    geom_bar(stat = "identity", position = "fill") +
    scale_fill_manual(values = c("Relapse" = "turquoise3",
                                 "Recrudescence" = "skyblue4",
                                 "Reinfection" = "magenta3")) +
  
    scale_x_discrete(expand = c(0, 0)) +
    scale_y_continuous(expand = c(0, 0)) +
    labs(x     = "Episode number", 
         y     = "Posterior probability", 
         fill = "") + 
    theme_bw() +
    facet_wrap(community ~ subject_id)
```
### By days since first episode during study period
```{r plot posterior probabilities with time, fig.height=8, fig.width=8}
marginal_summary %>% 
  left_join(analysis_data %>% select(subject_id, episode_number, community, visit_date, days_since_enrolment, days_since_last_episode) %>% distinct(), 
            by = c("subject_id", "episode_number")) %>% 
  group_by(subject_id) %>%
  arrange(desc(posterior_value), .by_group = TRUE) %>%
  mutate(subject_id = factor(subject_id, levels = unique(subject_id[order(posterior_value, decreasing = TRUE)]))) %>%
  ggplot(aes(x = days_since_enrolment, y = posterior_value, group = episode_number, fill = posterior_classification)) + 
  geom_bar(stat = "identity", position = "fill", width = 10) +
    scale_fill_manual(values = c("Relapse" = "turquoise3",
                                 "Recrudescence" = "skyblue4",
                                 "Reinfection" = "magenta3")) +
  
    scale_x_continuous(expand = c(0, 0)) +
    scale_y_continuous(expand = c(0, 0)) +
  labs(x     = "Days since first episode", 
       y     = "Posterior probability", 
       fill = "") + 
  theme_bw() +
  facet_wrap(~subject_id)
```
## Plot joint probabilities
```{r color palettes}
pal1 <- c(
  "C" = "turquoise3", 
  "I" = "magenta3", 
  "L" = "skyblue4")

pal2 <- c(
  "II" = "magenta3",
  "IL" = "darkorange2",
  "LI" = "goldenrod2",
  "LL" = "skyblue4",
  "CI" = "saddlebrown",
  "CL" = "turquoise4"
)

pal3 <- c(
  "LLL" = "skyblue4",
  "LIL" = "goldenrod3",
  "ILL" = "orangered3",
  "IIL" = "darkorange3",
  "LLI" = "orange3",
  "ILI" = "sienna4",
  "LII" = "peru",
  "III" = "magenta3"
)

combined_pal <- c(pal1, pal2, pal3)
```

```{r plotting joint fxn}
plotJointProb <- function(data, recurrence_n, prob_threshold, color_palette, ...){
  data %>% 
    filter(total_recurrences == recurrence_n, joint_probability > prob_threshold) %>% 
    group_by(subject_id) %>% 
    mutate(state_pair = fct_reorder(state_pair, joint_probability)) %>%
    ungroup() %>% 
    ggplot(aes(x = joint_probability, 
               y = subject_id, 
               fill = state_pair)) +
      geom_bar(position = "stack", stat = "identity") +
      scale_x_continuous(expand = c(0, 0)) +
      scale_y_discrete(expand = c(0, 0)) +
      scale_fill_manual(values = color_palette) +
      labs(x = "Joint probability estimate",
           y = "Participant",
           fill = "Classification state") +
      theme_bw() + 
    facet_wrap(~total_recurrences, scales = "free_y")
}
```

### By total number of recurrences
```{r plot joint by recurrences, fig.height = 7, fig.width = 12}
plot_joint1 <- plotJointProb(joint_summary, 1, 0, pal1) 
plot_joint2 <- plotJointProb(joint_summary, 2, 0, pal2) + guides(fill = guide_legend(ncol = 2)) 
plot_joint3 <- plotJointProb(joint_summary, 3, 0.001, pal3) + guides(fill = guide_legend(ncol = 3)) 

(plot_joint1 | plot_joint2) / (plot_joint3) +
  plot_annotation(subtitle = "I = Reinfection, L = Relapse, C = Recrudescence",
                  caption = expression(italic("For >2 recurrences, only probability estimates > 0.001 are shown"))) +
  plot_layout(heights = c(4, 0.5))
```
### By community and number of recurrences
```{r plot joint by treatment arm, fig.height = 8, fig.width = 12}
joint_summary %>% 
    left_join(analysis_data %>% select(subject_id, community) %>% distinct(),
              by = "subject_id") %>% 
    filter(joint_probability > 0.001) %>% 
    group_by(subject_id) %>% 
    mutate(state_pair = fct_reorder(state_pair, joint_probability)) %>%
    ungroup() %>% 
    ggplot(aes(x = joint_probability, 
               y = reorder(subject_id, total_recurrences), 
               fill = state_pair)) +
      geom_bar(position = "stack", stat = "identity") +
      scale_x_continuous(expand = c(0, 0)) +
      scale_y_discrete(expand = c(0, 0)) +
      scale_fill_manual(values = combined_pal) +
      labs(x = "Joint probability estimate",
           y = "Participant",
           fill = "Classification state",
           subtitle = "I = Reinfection, L = Relapse, C = Recrudescence",
           caption = expression(italic("Color coding: if > 1 recurrence, blue shades denote combinations of relapse and recrudescence (L and C), shades of orange/browns denote \ncombos of reinfection and relapse (I and L) and green shades denote combos of reinfection with recrudescence (I and C)"))) +
      theme_bw() + 
      facet_wrap(community~total_recurrences, scales = "free") +
      # facet_wrap(total_recurrences~treatment_arm, scales = "free") +
      theme(plot.caption = element_text(hjust = 0, vjust = -2)) 
```
### 'Confidence' in highest joint probability estimate
What is the joint probability estimate for the 'highest ranking' state pair classification? I.e. if >80% relatively high confidence? n=19
```{r plot joint probability confidence}
joint_summary %>% 
  filter(joint_probability > 0.001) %>% 
  group_by(subject_id) %>% 
  # get the highest prob for each subj and store ranking
  mutate(ranking = rank(-joint_probability, ties.method = "first")) %>% 
  arrange(subject_id, ranking) %>% 
  filter(ranking == 1) %>% 
  ggplot(aes(x = reorder(subject_id, -joint_probability),
             y = joint_probability,
             fill = factor(total_recurrences))) +
    geom_col() +
    geom_hline(yintercept = 0.8, color = "darkgray", linetype = "dashed") +
    labs(x = "Participant",
         y = "Joint probability estimate for highest 'ranking' state pair",
         fill = "Number of recurrent episodes",
         caption = expression(italic("Dashed line indicates an arbitrary threshold of 80%"))) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

How many state pairs make up the arbitrary 80% threshold?
```{r plot joint prob for pair rankings, fig.height=6, fig.width=9}
joint_summary %>% 
  filter(joint_probability > 0.001) %>% 
  group_by(subject_id) %>% 
  # Get the highest prob for each subject and store ranking
  mutate(ranking = rank(-joint_probability, ties.method = "first")) %>% 
  # reverse ranking order of ranking so we plot the highest rank at the bottom
  arrange(subject_id, desc(ranking)) %>%
  mutate(ranking = factor(ranking, levels = unique(ranking))) %>% 
  ggplot(aes(x = subject_id,
             y = joint_probability,
             fill = ranking)) +
    geom_bar(stat = "identity", position = "stack") +
    geom_hline(yintercept = 0.8, color = "darkgray", linetype = "dashed") +
    scale_fill_brewer(type = "qual", "Paired") + 
    # scale_fill_manual(values = colorRampPalette(brewer.pal(12, "Paired"))(14),
    #                   guide = guide_legend(reverse = T)) +
    scale_x_discrete(expand = c(0, 0)) +
    scale_y_continuous(expand = c(0, 0)) +
    labs(x = "Participant",
         y = "Joint probability estimate for state pair",
         fill = "Number of possible \nestimated state pairs",
         caption = expression(italic("Dashed line indicates an arbitrary threshold of 80%"))) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
    facet_wrap(~total_recurrences, scales = "free_x")
```

### Highest ranking classification by episode number
```{r create summary of top 3 rankings}
joint_top3_summary <- joint_summary %>% 
            select(-percentage) %>% 
            group_by(subject_id) %>% 
            # Get the highest prob for each subject and store ranking
            mutate(ranking = rank(-joint_probability, ties.method = "first")) %>% 
            arrange(subject_id, ranking) %>% 
            filter(ranking %in% c(1, 2, 3))
```

We will keep the 'highest ranking' joint probability estimate so we can compare the state classification for each episode with other metrics
```{r create summary of top ranking joint probability for each recurrence}
joint_recurrence_summary <- joint_top3_summary %>% 
  select(-total_recurrences) %>% 
  filter(ranking == 1) %>% 
  mutate(state_pair_split = strsplit(state_pair, "")) %>%  # Split state_pair into individual letters
  unnest(state_pair_split) %>%  # Unnest to create a new row for each letter
  group_by(subject_id) %>%
  mutate(
    episode_number = row_number() + 1,  # Create episode number starting from n+1 to reflect number of recurrence
    joint_probability = joint_probability,  # Keep the same joint_probability for all episodes\
    state_classification = case_when(state_pair_split == "L" ~ "Relapse",
                                     state_pair_split == "C" ~ "Recrudescence",
                                     state_pair_split == "I" ~ "Reinfection")) %>%
  select(subject_id, episode_number, state_classification, joint_probability)
```

```{r plot joint prob highest ranking by episode, fig.height=6, fig.width=8}
joint_recurrence_summary %>% 
  ggplot(aes(x = factor(episode_number), y = joint_probability, fill = state_classification)) +
    geom_col() +
    scale_fill_manual(values = c("Relapse" = "turquoise3",
                                 "Recrudescence" = "skyblue4",
                                 "Reinfection" = "magenta3")) +
    scale_x_discrete(expand = c(0, 0)) +
    scale_y_continuous(expand = c(0, 0)) +
    labs(x     = "Episode number", 
         y     = "Posterior joint probability", 
         fill = "") + 
    theme_bw() +
    facet_wrap(~subject_id)
```

## Comparison marginal vs joint 
```{r create posterior summary}
posterior_summary <- marginal_summary %>% 
  rename("marginal_posterior_probability" = "posterior_value",
         "marginal_classification" = "posterior_classification") %>% 
  left_join(joint_recurrence_summary %>% 
               rename("joint_classification" = "state_classification",
                      "joint_posterior_probability" = "joint_probability"),
             by = c("subject_id", "episode_number")) %>% 
  pivot_longer(cols = c(marginal_posterior_probability, 
                        joint_posterior_probability, 
                        marginal_classification, 
                        joint_classification),
    names_to = c("estimate_type", ".value"),
    names_pattern = "(marginal|joint)_(posterior_probability|classification)") %>% 
  distinct() 
```

```{r plot marginal with joint probs, fig.height = 6, fig.width=8}
posterior_summary %>% 
  group_by(subject_id, episode_number) %>%
  arrange(desc(posterior_probability), .by_group = TRUE) %>%
  mutate(classification = factor(classification, levels = unique(classification))) %>%
  ungroup() %>% 
  ggplot(aes(x = factor(episode_number), y = posterior_probability, group = episode_number)) + 
    geom_bar(data = . %>% filter(estimate_type == "marginal"),
             aes(fill = classification),
             stat = "identity", 
             position = "fill") +
    geom_segment(data = . %>% filter(estimate_type == "joint"),
               aes(x = as.numeric(factor(episode_number)) - 0.4,  
                   xend = as.numeric(factor(episode_number)) + 0.4,  
                   y = posterior_probability,  
                   yend = posterior_probability, 
                   color = classification),
               size = 1.5) +
    geom_point(data = . %>% filter(estimate_type == "joint"),
               aes(color = classification,
                   fill = classification),
               shape = 21) +
    scale_fill_manual(values = c("Relapse" = "turquoise3",
                                 "Recrudescence" = "skyblue4",
                                 "Reinfection" = "magenta3")) +
    scale_color_manual(values = c("Relapse" = "#006064",
                                 "Recrudescence" = "#3B4F75",
                                 "Reinfection" = "#8B008B")) +
    scale_x_discrete(expand = c(0, 0)) +
    scale_y_continuous(expand = c(0, 0)) +
    labs(x     = "Episode number", 
         y     = "Posterior probability", 
         fill = "Marginal classification",
         color = "Joint probability estimate \n and classification") + 
    theme_bw() +
    facet_wrap(~subject_id)
```

## Comparison to IBS and IBD estimates
```{r read pnh and ibd data}
peru_pnh <- read.csv(here("data/final", "peru_PNH_data.csv"))
ibd_r_data <- readRDS(here("data/final", "peru_all_meta.rds"))
ibd_rhat_data <- readRDS(here("data/final", "peru_sig_meta.rds"))
```

For direct comparison with the we can only look at the paired comparisons between baseline samples and 'follow-up' recurrent samples, but not other pw comparisons for a given participant. 
```{r filter datasets}
peru_ibd <- ibd_r_data %>% 
              # filter to include only paired samples
              filter(comparison_type == "paired") %>% 
              # merge with analysis_data to get patient_name and 'day' info
              left_join(analysis_data %>% 
                          select(sample_id, subject_id, 
                                 day = days_since_enrolment) %>% 
                          distinct(),
                        by = c("sampleid1" = "sample_id")) %>% 
              rename("subject_id1" = "subject_id",
                     "day_id1" = "day") %>% 
              # do the same for sample id 2 in pair - merge with analysis_data to get patient_name and 'day' info
              left_join(analysis_data %>% 
                          select(sample_id, subject_id, 
                                 day = days_since_enrolment) %>% 
                          distinct(),
                        by = c("sampleid2" = "sample_id")) %>% 
              rename("day_id2" = "day") %>% 
              select(-subject_id1) %>% 
              # filter to include only pw comparisons to first episode in study period
              filter(day_id2 == "0") %>% 
              group_by(subject_id) %>%  
              arrange(date1) %>%  
              # create episode_number accounting for baseline as episode 1
              mutate(episode_number = row_number() + 1) %>% 
              ungroup()
```

```{r classification summary}
classification_summary <- indiv_posteriors_marginal %>% 
  rename("posterior_probability_recrudescence" = "Posterior_marginal_prob_C",
         "posterior_probability_relapse" = "Posterior_marginal_prob_L",
         "posterior_probability_reinfection" = "Posterior_marginal_prob_I") %>% 
  # merge with epi variables
  left_join(analysis_data %>% select(subject_id, sample_id, episode_number, community, visit_date, days_since_enrolment, days_since_last_episode) %>% distinct(), 
            by = c("subject_id", "episode_number")) %>% 
  # merge with PNH data
  left_join(peru_pnh,
            by = c("subject_id" = "Patient", 
                   "episode_number" = "recurrence")) %>% 
  rename("prop_same_haps_1minusPNH" = "Prop_het", 
         "pnh_range" = "identity", 
         "classification" = "Classification") %>% 
  # create bins for 1-PNH (IBS metric)
  mutate(ibs_range = cut(prop_same_haps_1minusPNH,  
                         breaks = c(0, 0.25, 0.5, 0.75, 1), 
                         include.lowest = TRUE, 
                         labels = c("0-0.25", "0.25-0.5", "0.5-0.75", "0.75-1"))) %>% 
  # merge with IBD data
  left_join(peru_ibd,
            by = c("subject_id", "episode_number")) %>% 
  # create bins for IBD
  mutate(ibd_range = cut(estimate,  
                         breaks = c(0, 0.25, 0.5, 0.75, 1), 
                         include.lowest = TRUE, 
                         labels = c("0-0.25", "0.25-0.5", "0.5-0.75", "0.75-1"))) %>% 
  # merge with joint probabilities
  left_join(joint_recurrence_summary %>% 
              rename("joint_classification" = "state_classification"),
            by = c("subject_id", "episode_number"))

classification_summary
```

### Comparison with IBS
```{r classification comparison with IBS, eval=F}
classification_summary %>% 
  select(subject_id, sample_id, episode_number, visit_date, days_since_enrolment, days_since_last_episode, starts_with("posterior_"), prop_same_haps_1minusPNH, pnh_range, ibs_range, classification) %>% 
  pivot_longer(cols = (starts_with("posterior")), 
               names_to = "posterior_type", 
               values_to = "posterior_value") %>% 
  mutate(posterior_range = cut(posterior_value, 
                               breaks = c(0, 0.25, 0.5, 0.75, 1), 
                               include.lowest = TRUE, 
                               labels = c("0-0.25", "0.25-0.5", "0.5-0.75", "0.75-1"))) %>% 
  count(ibs_range, classification, posterior_type, posterior_range)
```

### Plot IBS range vs probabilistic classification (marginal probability)
```{r classification IBS comparison plot v marginal, fig.height=10, fig.width=8}
classification_summary %>% 
  select(subject_id, sample_id, pair, episode_number, visit_date, days_since_enrolment, days_since_last_episode, starts_with("posterior_"), prop_same_haps_1minusPNH, pnh_range, ibs_range, classification) %>% 
  pivot_longer(cols = (starts_with("posterior")), 
               names_to = "posterior_type", 
               values_to = "posterior_value") %>% 
  mutate(posterior_range = cut(posterior_value, 
                               breaks = c(0, 0.25, 0.5, 0.75, 1), 
                               include.lowest = TRUE, 
                               labels = c("0-0.25", "0.25-0.5", "0.5-0.75", "0.75-1"))) %>%  
  ggplot(aes(x = reorder(pair, episode_number), y = posterior_value, group = episode_number, fill = posterior_type)) +
    geom_bar(stat = "identity", position = "fill") +
    scale_fill_manual(values = c("turquoise3", "magenta3", "skyblue4"),
                       labels = c("posterior_probability_relapse" = "Relapse",
                                  "posterior_probability_recrudescence" = "Recrudescence",
                                  "posterior_probability_reinfection" = "Reinfection")) +
    scale_x_discrete(expand = c(0, 0)) +
    scale_y_continuous(expand = c(0, 0)) +
    labs(x     = "Sample", 
         y     = "Posterior probability", 
         fill = "") + 
    theme_bw() +
    facet_wrap(~ibs_range, scales = "free_x") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```
### Plot IBS range vs probabilistic classification (joint probability, highest ranking)
```{r classification IBS comparison plot v joint, fig.height=10, fig.width=8}
classification_summary %>% 
  select(subject_id, sample_id, pair, episode_number, visit_date, days_since_enrolment, days_since_last_episode, starts_with("joint"), prop_same_haps_1minusPNH, pnh_range, ibs_range, classification) %>% 
  ggplot(aes(x = reorder(pair, episode_number), y = joint_probability, group = episode_number, fill = joint_classification)) +
    geom_bar(stat = "identity") +
    scale_fill_manual(values = c("Relapse" = "turquoise3",
                                 "Recrudescence" = "skyblue4",
                                 "Reinfection" = "magenta3")) +
    scale_x_discrete(expand = c(0, 0)) +
    scale_y_continuous(expand = c(0, 0)) +
    labs(x     = "Sample pair", 
         y     = "Posterior joint probability (highest ranking classification)", 
         fill = "") + 
    theme_bw() +
    facet_wrap(~ibs_range, scales = "free_x") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

### Comparison with IBD
```{r classification comparison with IBD, eval=F}
classification_summary %>% 
  select(subject_id, sample_id, episode_number, visit_date, days_since_enrolment, days_since_last_episode, starts_with("posterior_"), ibd_range, CI_lower, CI_upper, sig_est) %>% 
  pivot_longer(cols = (starts_with("posterior")), 
               names_to = "posterior_type", 
               values_to = "posterior_value") %>% 
  mutate(posterior_range = cut(posterior_value, 
                               breaks = c(0, 0.25, 0.5, 0.75, 1), 
                               include.lowest = TRUE, 
                               labels = c("0-0.25", "0.25-0.5", "0.5-0.75", "0.75-1"))) %>% 
  count(ibd_range, sig_est, posterior_type, posterior_range)
```

### For reference - all pairwise comparisons
```{r plot estimates and CI all pw comparisons, fig.height=16, fig.width=10}
ibd_r_data %>% 
  filter(comparison_type == "paired") %>% 
  ggplot(aes(x = pair, y = estimate, group = patientid1, color = sig_est)) +
    geom_point(size = 3) +
    geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper)) +
    scale_color_manual(values = c("not significant" = "darkgrey", 
                                  "significant" = "indianred3")) +
    facet_wrap(~patientid1, scales = "free_x") +
    theme_bw() +    
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

### Plot IBD estimate and CI for paired samples compared to baseline for each participant
```{r plot estimates and CI paired samples, fig.height=12, fig.width=8}
classification_summary %>% 
  ggplot(aes(x = reorder(sampleid1, days_since_enrolment), y = estimate, group = subject_id, color = sig_est)) +
    geom_point(size = 3) +
    geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper)) +
    scale_color_manual(values = c("not significant" = "darkgrey", 
                                  "significant" = "indianred3")) +
    labs(x = "Sample",
         y = "IBD",
         color = "Significance") +
    facet_wrap(~subject_id, scales = "free_x") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

```{r plot estimates and CI not paired samples, eval=F}
ibd_r_data %>% 
  filter(comparison_type == "not paired") %>% 
  ggplot(aes(x = pair, y = estimate, group = patientid1, color = sig_est)) +
    geom_point(size = 3) +
    geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper)) +
    scale_color_manual(values = c("not significant" = "darkgrey", 
                                  "significant" = "indianred3")) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

### Plot IBD estimate and CI for paired samples compared to baseline for each participant, based on days since last episode
Perhaps reinfections are more likely to occur after more time?
```{r plot estimates and CI paired samples v days, fig.height=12, fig.width=8}
classification_summary %>% 
  ggplot(aes(x = days_since_last_episode, y = estimate, group = subject_id, color = sig_est)) +
    geom_point(size = 3) +
    geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper)) +
    scale_color_manual(values = c("not significant" = "darkgrey", 
                                  "significant" = "indianred3")) +
    labs(x = "Days since last episode",
         y = "IBD",
         fill = "Significance") +
    facet_wrap(~subject_id, scales = "free_x") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

### Plot IBD range vs probabilistic classification (marginal)
```{r classification IBD comparison plot, fig.height=10, fig.width=8}
classification_summary %>% 
  select(subject_id, sample_id, pair, episode_number, visit_date, days_since_enrolment, days_since_last_episode, starts_with("posterior_"), ibd_range, CI_lower, CI_upper, sig_est) %>% 
  pivot_longer(cols = (starts_with("posterior")), 
               names_to = "posterior_type", 
               values_to = "posterior_value") %>% 
  mutate(posterior_range = cut(posterior_value, 
                               breaks = c(0, 0.25, 0.5, 0.75, 1), 
                               include.lowest = TRUE, 
                               labels = c("0-0.25", "0.25-0.5", "0.5-0.75", "0.75-1"))) %>%  
  ggplot(aes(x = reorder(pair, episode_number), y = posterior_value, group = episode_number, fill = posterior_type)) +
    geom_bar(stat = "identity", position = "fill") +
    scale_fill_manual(values = c("turquoise3", "magenta3", "skyblue4"),
                       labels = c("posterior_probability_relapse" = "Relapse",
                                  "posterior_probability_recrudescence" = "Recrudescence",
                                  "posterior_probability_reinfection" = "Reinfection")) +
    labs(x     = "Sample pair", 
         y     = "Posterior probability", 
         fill = "") + 
    theme_bw() +
    facet_wrap(~ibd_range, scales = "free_x") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

### Plot IBD range vs probabilistic classification (joint probability, highest ranking)
```{r classification IBD comparison plot v joint, fig.height=10, fig.width=8}
classification_summary %>% 
  select(subject_id, sample_id, pair, episode_number, visit_date, days_since_enrolment, days_since_last_episode, starts_with("joint"), ibd_range, CI_lower, CI_upper, sig_est) %>% 
  ggplot(aes(x = reorder(pair, episode_number), y = joint_probability, group = episode_number, fill = joint_classification)) +
    geom_bar(stat = "identity") +
    scale_fill_manual(values = c("Relapse" = "turquoise3",
                                 "Recrudescence" = "skyblue4",
                                 "Reinfection" = "magenta3")) +
    scale_x_discrete(expand = c(0, 0)) +
    scale_y_continuous(expand = c(0, 0)) +
    labs(x     = "Sample pair", 
         y     = "Posterior joint probability (highest ranking classification)", 
         fill = "") + 
    theme_bw() +
    facet_wrap(~ibd_range, scales = "free_x") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

## Confusion matrices: PNH vs probabilistic classification

### Classification by 1-PNH (IBS) range
```{r data for confusion matrix ibs range}
confusion_data_ibsrange <- classification_summary %>% 
  select(subject_id, sample_id, episode_number, visit_date, days_since_enrolment, days_since_last_episode, starts_with("posterior_"), starts_with("joint"), prop_same_haps_1minusPNH, pnh_range, ibs_range, 
         # renaming for clarity
         pnh_classification = classification) %>% 
  pivot_longer(cols = (starts_with("posterior")), 
               names_to = "marginal_classification", 
               values_to = "marginal_probability") %>% 
  mutate(marginal_classification = case_when(marginal_classification == "posterior_probability_recrudescence" ~ "Recrudescence",
                                             marginal_classification == "posterior_probability_relapse" ~ "Relapse",
                                             marginal_classification == "posterior_probability_reinfection" ~ "Reinfection"),
         marginal_classification = factor(marginal_classification, 
                                          levels = c("Relapse", "Recrudescence", "Reinfection")),
         marginal_probability_range = cut(marginal_probability, 
                                          breaks = c(0, 0.25, 0.5, 0.75, 1), 
                                          include.lowest = TRUE, 
                                          labels = c("0-0.25", "0.25-0.5", "0.5-0.75", "0.75-1")),
         joint_probability_range = cut(joint_probability, 
                                       breaks = c(0, 0.25, 0.5, 0.75, 1), 
                                       include.lowest = TRUE, 
                                       labels = c("0-0.25", "0.25-0.5", "0.5-0.75", "0.75-1"))) %>% 
  select(subject_id, episode_number, days_since_enrolment, ibs_range, pnh_classification,  marginal_classification, marginal_probability, marginal_probability_range, joint_classification, joint_probability, joint_probability_range) %>% 
  # now only keep the highest 'ranking' marginal classification
  group_by(subject_id, episode_number) %>% 
  slice_max(order_by = marginal_probability, n = 1, with_ties = F) %>% 
  ungroup() %>% 
  group_by(ibs_range, marginal_classification, joint_classification) %>%
  tally() %>%
  dcast(ibs_range ~ marginal_classification + joint_classification, value.var = "n", fill = 0)
```

```{r plot confusion matrix ibs range}
confusion_data_ibsrange %>% 
  melt(id.vars = "ibs_range") %>% 
  ggplot(aes(ibs_range, variable, fill = value)) +
    geom_tile() +
    scale_fill_gradient(low = "white", high = "red") +
    geom_text(aes(label = value)) +
    theme_minimal() +
    labs(x = "IBS range (1-PNH)",
         y = "Marginal vs joint classification",
         fill = "Count")
```

### Classification comparison
```{r data for confusion matrix pnh state}
# Create a confusion matrix for each pair
confusion_data_pnhstate <- 
  classification_summary %>% 
  select(subject_id, sample_id, episode_number, visit_date, days_since_enrolment, days_since_last_episode, starts_with("posterior_"), starts_with("joint"), prop_same_haps_1minusPNH, pnh_range, ibs_range, 
         # renaming for clarity
         pnh_classification = classification) %>% 
  pivot_longer(cols = (starts_with("posterior")), 
               names_to = "marginal_classification", 
               values_to = "marginal_probability") %>% 
  mutate(marginal_classification = case_when(marginal_classification == "posterior_probability_recrudescence" ~ "Recrudescence",
                                             marginal_classification == "posterior_probability_relapse" ~ "Relapse",
                                             marginal_classification == "posterior_probability_reinfection" ~ "Reinfection"),
         marginal_classification = factor(marginal_classification, 
                                          levels = c("Relapse", "Recrudescence", "Reinfection")),
         marginal_probability_range = cut(marginal_probability, 
                                          breaks = c(0, 0.25, 0.5, 0.75, 1), 
                                          include.lowest = TRUE, 
                                          labels = c("0-0.25", "0.25-0.5", "0.5-0.75", "0.75-1")),
         joint_probability_range = cut(joint_probability, 
                                       breaks = c(0, 0.25, 0.5, 0.75, 1), 
                                       include.lowest = TRUE, 
                                       labels = c("0-0.25", "0.25-0.5", "0.5-0.75", "0.75-1")),
         # pnh_state = case_when(pnh_classification == "Heterologous" ~ "Reinfection",
         #                       pnh_classification == "Homologous" ~ "Relapse or Recrudescence")
         pnh_state = case_when(ibs_range == "0-0.25" | ibs_range == "0.25-0.5" ~ "Reinfection",
                               (ibs_range == "0.5-0.75" | ibs_range == "0.75-1") ~ "Relapse or Recrudescence")
         ) %>% 
  select(subject_id, episode_number, days_since_enrolment, ibs_range, pnh_classification, pnh_state, marginal_classification, marginal_probability, marginal_probability_range, joint_classification, joint_probability, joint_probability_range) %>% 
  # now only keep the highest 'ranking' marginal classification
  group_by(subject_id, episode_number) %>% 
  slice_max(order_by = marginal_probability, n = 1, with_ties = F) %>% 
  ungroup() %>% 
  group_by(pnh_state, marginal_classification, joint_classification) %>%
  tally() %>%
  dcast(pnh_state ~ marginal_classification + joint_classification, value.var = "n", fill = 0)
```

```{r plot confusion matrix pnh state}
confusion_data_pnhstate %>% 
  melt(id.vars = "pnh_state") %>% 
  ggplot(aes(pnh_state, variable, fill = value)) +
    geom_tile() +
    scale_fill_gradient(low = "white", high = "red") +
    geom_text(aes(label = value)) +
    theme_minimal() +
    labs(x = "PNH classification",
         y = "Marginal vs joint classification",
         fill = "Count")
```
